---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parttree

<!-- badges: start -->
[![R-CMD-check](https://github.com/grantmcdermott/parttree/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/grantmcdermott/parttree/actions/workflows/R-CMD-check.yaml)
[![Docs](https://img.shields.io/badge/docs-homepage-blue.svg)](https://grantmcdermott.com/parttree/index.html)
<!-- badges: end -->

Visualize simple 2-D decision tree partitions in R. The **parttree**
package is optimised to work with [**ggplot2**](https://ggplot2.tidyverse.org/),
although it can be used to visualize tree partitions with base R graphics too.


[`Installation`](#installation) | [`Examples`](#examples) | 
[`Supported model classes`](#supported-model-classes) | 
[`Plot orientation`](#plot-orientation) | [`Base graphics`](#base-graphics)

## Installation

This package is not yet on CRAN, but can be installed from [GitHub](https://github.com/) 
with:

``` r
# install.packages("remotes")
remotes::install_github("grantmcdermott/parttree")
```
## Examples

The main function that users will interact with is `geom_parttree()`. Here's a
simple example using the [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/) 
dataset.

```{r penguin_plot}
library("palmerpenguins") ## For 'penguins' dataset
library("rpart")          ## For fitting decisions trees
library("parttree")       ## This package (will automatically load ggplot2 too)

## First construct a scatterplot of the raw penguin data
p = ggplot(data = penguins, aes(x = flipper_length_mm, y = bill_length_mm)) +
    geom_point(aes(col = species)) +
    theme_minimal()

## Fit a decision tree using the same variables as the above plot
tree = rpart(species ~ flipper_length_mm + bill_length_mm, data = penguins)

## Visualise the tree partitions by adding it via geom_parttree()
p +  
  geom_parttree(data = tree, aes(fill=species), alpha = 0.1) +
  labs(caption = "Note: Points denote observed data. Shaded regions denote tree predictions.")
```

Trees with continuous independent variables are also supported. However, I 
recommend adjusting the plot fill aesthetic, since your tree will likely 
partition the data into intervals that don't match up exactly with the raw data.

```{r penguin_plot2}
tree2 = rpart(body_mass_g ~ flipper_length_mm + bill_length_mm, data=penguins)

p2 = 
  ggplot(data = penguins, aes(x = flipper_length_mm, y = bill_length_mm)) +
  geom_parttree(data = tree2, aes(fill=body_mass_g), alpha = 0.3) +
  geom_point(aes(col = body_mass_g)) +
  theme_minimal()
  
## Legend scales don't quite match (try it yourself)
# p2

## Better to scale fill to the original data. This does the job but is still 
## kind of hard to make out (again, try yourself)
# p2 + 
#   scale_fill_continuous(limits = range(penguins$body_mass_g, na.rm = TRUE))

## Best and simplest just to specify colour and fill aesthetics together in
## single palette
p2 + 
  scale_colour_viridis_c(aesthetics = c('colour', 'fill'))
```

## Supported model classes

Currently, the package works with decision trees created by the 
[**rpart**](https://CRAN.R-project.org/web/package=rpart) and
[**partykit**](https://CRAN.R-project.org/web/package=partykit) packages. 
Moreover, it supports other front-end modes that call `rpart::rpart()` as 
the underlying engine; in particular the
[**tidymodels**](https://www.tidymodels.org/) ([parsnip](https://parsnip.tidymodels.org/) 
or [workflows](https://workflows.tidymodels.org/)) and 
[**mlr3**](https://mlr3.mlr-org.com/) packages. Here's an example with parsnip.

```{r titanic_plot}
library("parsnip")
library("titanic") ## Just for a different data set
set.seed(123) ## For consistent jitter

titanic_train$Survived = as.factor(titanic_train$Survived)

## Build our tree using parsnip (but with rpart as the model engine)
ti_tree =
  decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification") %>%
  fit(Survived ~ Pclass + Age, data = titanic_train)

## Plot the data and model partitions
titanic_train %>%
  ggplot(aes(x=Pclass, y=Age)) +
  geom_jitter(aes(col=Survived), alpha=0.7) +
  geom_parttree(data = ti_tree, aes(fill=Survived), alpha = 0.1) +
  theme_minimal()
```

## Plot orientation

Underneath the hood, `geom_parttree()` is calling the companion `parttree()` 
function, which coerces the **rpart** tree object into a data frame that is 
easily understood by **ggplot2**. For example, consider again our first "tree" 
model from earlier. Here's the print output of the raw model.

```{r tree}
tree
```

And here's what we get after we feed it to `parttree()`.

```{r tree_parted}
parttree(tree)
```

Again, the resulting data frame is designed to be amenable to a **ggplot2** geom
layer, with columns like `xmin`, `xmax`, etc. specifying aesthetics that 
**ggplot2** recognises. (Fun fact: `geom_parttree()` is really just a thin 
wrapper around `geom_rect()`.) The goal of the package is to abstract away these 
kinds of details 
from the user, so we can just specify `geom_parttree()` &mdash; with a valid
tree object as the data input &mdash; and be done with it. However, while this 
generally works well, it can sometimes lead to unexpected behaviour in terms of 
plot orientation. That's because it's hard to guess ahead of time what the user 
will specify as the x and y variables (i.e. axes) in their other plot layers. To 
see what I mean, let's redo our penguin plot from earlier, but this time switch 
the axes in the main `ggplot()` call.

```{r tree_plot_mismatch}
## First, redo our first plot but this time switch the x and y variables
p3 = 
  ggplot(
    data = penguins, 
    aes(x = bill_length_mm, y = flipper_length_mm) ## Switched!
    ) +
  geom_point(aes(col = species)) +  
  theme_minimal()

## Add on our tree (and some preemptive titling..)
p3 +
  geom_parttree(data = tree, aes(fill = species), alpha = 0.1) +
  labs(
    title = "Oops!",
    subtitle = "Looks like a mismatch between our x and y axes..."
    )
```

As was the case here, this kind of orientation mismatch is normally (hopefully) 
pretty easy to recognize. To fix, we can use the `flipaxes = TRUE` argument to 
flip the orientation of the `geom_parttree` layer.

```{r tree_plot_flip}
p3 +
  geom_parttree(
    data = tree, aes(fill = species), alpha = 0.1,
    flipaxes = TRUE  ## Flip the orientation
    ) +
  labs(title = "That's better")
```

## Base graphics

While the package has been primarily designed to work with **ggplot2**, the
`parttree()` infrastructure can also be used to generate plots with base
graphics. Here, the `ctree()` function from **partykit** is used for fitting
the tree.

```{r ctree_base_graphics}
library("partykit")

## CTree and corresponding partition
ct = ctree(species ~ flipper_length_mm + bill_length_mm, data = penguins)
pt = parttree(ct)

## Color palette
pal = palette.colors(4, "R4")[-1]

## Maximum/minimum for plotting range as rect() does not handle Inf well
m = 1000

## scatter plot() with added rect()
plot(bill_length_mm ~ flipper_length_mm, data = penguins, col = pal[species], pch = 19)
rect(pmax(-m, pt$xmin), pmax(-m, pt$ymin), pmin(m, pt$xmax), pmin(m, pt$ymax),
  col = adjustcolor(pal, alpha.f = 0.1)[pt$species])
```
