---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parttree

<!-- badges: start -->
<!-- badges: end -->

A set of simple functions for visualizing decision tree partitions in R with
[**ggplot2**](https://ggplot2.tidyverse.org/).

## Installation

This package is not yet on CRAN, but can be installed from [GitHub](https://github.com/) 
with:

``` r
# install.packages("remotes")
remotes::install_github("grantmcdermott/parttree")
```
## Example

```{r iris_plot}
library(parttree)
library(rpart)
library(ggplot2)

iris_tree = rpart(Species ~ Sepal.Width + Petal.Width, data=iris)

## Let's construct a scatterplot of the original iris data
p = ggplot(data = iris, aes(x=Petal.Width, y=Sepal.Width)) +
  geom_point(aes(col=Species))

## We now add the partitions with geom_parttree()
p +  
  geom_parttree(data = iris_tree, aes(fill=Species), alpha = 0.1) +
  labs(caption = "Note: Points denote observed data. Shaded regions denote tree predictions.") 
```

The main goal of the package is to visualize decision tree objects (e.g. created 
with [**rpart**](https://cran.r-project.org/web/packages/rpart/index.html)) into 
a data frame, which is amenable to visualizing as a partition plot. Here's a 
simple example using everyone's favourite `iris` dataset.

## Limitations and caveats

### Supported model classes

Currently, the function only works with decision trees created by the 
[**rpart**](https://cran.r-project.org/web/packages/rpart/index.html) package. 
However, it does support other packages and modes that call `rpart::rpart()` as 
the underlying engine. Here's an example using the 
[**parsnip**](https://tidymodels.github.io/parsnip/) package. 

```{r titanic_plot}
library(parsnip)
library(titanic) ## Just for a different data set
set.seed(123) ## For consistent jitter

titanic_train$Survived = as.factor(titanic_train$Survived)

## Build our tree using parsnip (but with rpart as the model engine)
ti_tree =
  decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification") %>%
  fit(Survived ~ Pclass + Age, data = titanic_train)

## Plot the data and model partitions
titanic_train %>%
  ggplot(aes(x=Pclass, y=Age)) +
  geom_jitter(aes(col=Survived), alpha=0.7) +
  geom_parttree(data = ti_tree, aes(fill=Survived), alpha = 0.1) +
  theme_minimal()
```

### Orientation

Underneath the hood, `geom_partree()` is calling the companion `parttree()` 
function, which coerces the **rpart** tree object into a data frame that is 
easily understood by **ggplot2**. For example, consider our "ti_tree" model from
above.

```{r titanic_parted}
parttree(ti_tree)
```

Again, the resulting data frame is designed to amenable to a **ggplot2** geom
layer, with columns like `xmin`, `xmax`, etc. specifying aesthetics that 
**ggplot2** recognises. (Fun fact: `geom_partree()` is really just a thin 
wrapper around `geom_rect()`). The goal is to abstract these kinds of details 
away from the user so we can just specify `geom_partree()` &mdash with a valid
tree object as the data input &mdash; and be done with it. However, while this 
generally works fine, it can sometimes lead to unexpected behaviour in terms of 
plot orientation. That's because it's hard to guess ahead of time what the user 
will specify as the *x* and *y* axes/variables in their other plot layers. To 
see what I mean, let's redo our titanic plot from earlier, but this time switch 
the axes in the main *ggplot()* call.

```{r titanic_plot_rot}
titanic_train %>%
  ggplot(aes(x=Age, y=Pclass)) + ## Changed!
  geom_jitter(aes(col=Survived), alpha=0.7) +
  geom_parttree(data = ti_tree, aes(fill=Survived), alpha = 0.1) +
  theme_minimal() +
  labs(title = "Oops!")
```

Now, normally this kind of orientation mismatch should hopefully be 
pretty easy to identify. But its admittedly annoying. I'll try to add better 
support for catching/avoiding these kinds of errors in a future update, but as 
of the moment *caveat emptor*.
