---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parttree

<!-- badges: start -->
<!-- badges: end -->

A set of simple functions for visualizing decision tree partitions in R.

## Installation

This package is not yet on CRAN, but can be installed from [GitHub](https://github.com/) 
with:

``` r
# install.packages("remotes")
remotes::install_github("grantmcdermott/parttree")
```
## Examples

The main goal of the package is to convert decision tree objects (e.g. created 
with [**rpart**](https://cran.r-project.org/web/packages/rpart/index.html)) into 
a data frame, which is amenable to visualizing as a partition plot. Here's a 
simple example using everyone's favourite `iris` dataset.

```{r example_rpart}
library(rpart)
library(parttree)

tree_mod = rpart(Species ~ Sepal.Width + Petal.Width, data=iris)
tree_mod

## Let's convert it to a partition data frame
tree_mod_parted = parttree(tree_mod)
tree_mod_parted
```
Again, the resulting data frame is designed to be compatible with 
[**ggplot2**](https://ggplot2.tidyverse.org/) (in particular, `geom_rect()`). 
This makes it convenient for visualizing partitions in conjunction with the 
original data.

```{r iris_plot}
library(ggplot2)

ggplot() +
  geom_point(data = iris, aes(x=Petal.Width, y=Sepal.Width, col=Species)) +
  geom_rect(
    data = tree_mod_parted, alpha = 0.1, col = "black",
    aes(xmin=xmin, ymin=ymin, xmax=xmax, ymax=ymax, fill=Species)
    ) +
  labs(caption = "Note: Points denote observed data. Shaded regions denote tree predictions.")
```

Currently, the function only works with decision trees created by the 
[**rpart**](https://cran.r-project.org/web/packages/rpart/index.html) package. 
However, it does support other packages and modes that call `rpart::rpart()` as 
the underlying engine. Here's an example using the 
[**parsnip**](https://tidymodels.github.io/parsnip/) package. 

```{r titanic_plot}
library(titanic) ## Just for a different data set
library(parsnip) ## Will use rpart engine
set.seed(123) ## for consistent jitter

titanic_train$Survived = as.factor(titanic_train$Survived)

## Build our tree using parsnip (but with rpart as the model engine)
ti_tree =
  decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification") %>%
  fit(Survived ~ Pclass + Age, data = titanic_train)

## Plot the data and model partitions
ggplot() +
  geom_jitter(
    data = titanic_train,
    aes(x=Pclass, y=Age, col=Survived), alpha=0.7
    ) +
  geom_rect(
    data = parttree(ti_tree), alpha = 0.1, col = "black",
    aes(xmin=xmin, ymin=ymin, xmax=xmax, ymax=ymax, fill=Survived)
    ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

