---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parttree

<!-- badges: start -->
<!-- badges: end -->

A set of simple functions for visualizing decision tree partitions in R.

## Installation

This package is not yet on CRAN, but can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("grantmcdermott/parttree")
```
## Examples

```{r example_rpart}
library(rpart)
library(parttree)

d = parttree(rpart(Species ~ Sepal.Width + Petal.Width, data=iris))
d
```

The return object is designed to be compatible with **ggplot2** (in particular, `geom_rect`). This makes it convenient for visualizing partitions in conjunction with the original data.

```{r iris_plot}
library(ggplot2)

ggplot() +
  geom_point(data = iris, aes(x=Petal.Width, y=Sepal.Width, col=Species)) +
  geom_rect(
    data = d, alpha = 0.1, col = "black",
    aes(xmin=xmin, ymin=ymin, xmax=xmax, ymax=ymax, fill=Species)
    ) 
```

Currently, the function only works with decision trees created by the **rpart** package. However, it does support other packages and modes that call `rpart()` as the underlying engine. Here's an example using the **parsnip** package. 

```{r titanic_plot}
library(titanic) ## Just for a different data set
library(parsnip)

titanic_train$Survived = as.factor(titanic_train$Survived)

## Build our tree using parsnip (but with rpart as the model engine)
ti_tree =
  decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification") %>%
  fit(Survived ~ Pclass + Age, data = titanic_train)

## Plot the data and model partitions
ggplot() +
  geom_jitter(
    data = titanic_train,
    aes(x=Pclass, y=Age, col=Survived), alpha=0.7
    ) +
  geom_rect(
    data = parttree(ti_tree), alpha = 0.1, col = "black",
    aes(xmin=xmin, ymin=ymin, xmax=xmax, ymax=ymax, fill=Survived)
    ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

